Modeling workflow: NLA N2O survey data

### 2024-03-18 scripts/NLA_N2O_models.Rmd
Updated modeling documentation

---
title: "Modeling workflow: NLA N2O survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

@@

# Load model and data files for knitting (if they already exist)
f1 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda"
f2 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda"
f3 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_test.rda"
f4 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda"
f5 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda"
f6 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/samp_props.rda"

f7 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod1.rda"
f8 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod2.rda"
f9 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod3.rda"
f10 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod4.rda"
f11 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod5.rda"
f12 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_mod6.rda"
f13 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/all_predictions_b.rda"
f14 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda"
f15 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda"
f16 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_survey_ests.rda"
f17 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/n2o_survey_ests_size.rda"
f18 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sat_survey_ests.rda"
f19 <- "C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sat_survey_ests_size.rda"

if(file.exists(f1))
  load(f1)
if(file.exists(f2))
  load(f2)
if(file.exists(f3))
  load(f3)
if(file.exists(f4))
  load(f4)
if(file.exists(f5))
  load(f5)
if(file.exists(f6))
  load(f6)
if(file.exists(f7))
  load(f7)
if(file.exists(f8))
  load(f8)
if(file.exists(f9))
  load(f9)
if(file.exists(f10))
  load(f10)
if(file.exists(f11))
  load(f11)
if(file.exists(f12))
  load(f12)
if(file.exists(f13))
  load(f13)
if(file.exists(f14))
  load(f14)
if(file.exists(f15))
  load(f15)
if(file.exists(f16))
  load(f16)
if(file.exists(f17))
  load(f17)
if(file.exists(f18))
  load(f18)
if(file.exists(f19))
  load(f19)
```

# Background and Objectives
This document describes the modeling workflow and results from estimating dissolved and equilibrium N2O gas concentrations and saturation ratios for all freshwater waterbodies (natural and man-made lakes and reservoirs) in the lower 48 US states larger than 1 ha. Data for these estimates came from the 2017 Nation Lakes Assessment (NLA) survey, wherein a subset of waterbodies were sampled from the broader population of interest (POI) according to a spatially balanced, stratified, and unequal probability design. Stratification was among categories of lake size (surface area in hectares), WSA9 ecoregion, and US state (excluding AK and HI). In order to better capture variability among the wide range of lake sizes in the POI, larger lakes were intentionally over-sampled relative to smaller lakes.

To make inferences back to the POI from this sample, estimates needed to be adjusted in a manner that addressed potential biases related to the relative representation of waterbody characteristics in the sample compared to the POI, specifically with respect to the survey design variables. It can be helpful to define the complete data likelihood (CDL) when considering how to address biases that may arise from sampling designs [@Link_Barker_2010; @Zachmann_etal_2022]. For the 2017 NLA survey, the complete data would be considered all US lakes and reservoirs in the lower 48 states larger than 1 hectare (_i.e._, the POI). The survey samples, by comparison, could then be considered a subset of that complete data with a known pattern of "missingess" assigned by the survey. In this case, waterbodies in the POI were missing from the sample not completely at random, but according to a probability assigned by the design parameters. This type of missingness is considered not ignorable when making inferences from a sample to a POI [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022]. However, _conditional_ on the design variables, the sample could be considered a simple random sample (SRS) from the POI [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022].

A simple, model-based strategy for adjusting a sample to a POI is to include the survey design variables as predictors in a regression model [@Gelman_etal_2020, Ch. 17]. After fitting the regression, the parameter estimates can be used to make adjustments according to the distribution of the design variables in the POI in a second step referred to as "poststratification" [@Gelman_etal_2020, Ch. 17]. A particularly popular version of this approach uses multilevel regression models and is referred to as multilevel regression with poststratification, or MRP [@Park_etal_2004; @Gelman_Little_1997; @Gelman_etal_2020, Ch. 17; @Kennedy_Gelman_2021]. Multilevel regression models are often recommended because they can provide regularized estimates along the design groupings, which can improve out-of-sample inferences [_e.g._, @Gelman_Little_1997; @Kennedy_Gelman_2021]. Estimates for group levels that may be missing from the sample, but are part of the population of interest, are also straightforward using the multilevel approach [@Gelman_etal_2020 Ch. 17; @McElreath_2020]. For a recent applied example and helpful tutorial employing this approach, see @Kennedy_Gelman_2021. For an example of a recent application in the context of national surveys of environmental resources, see @Zachmann_etal_2022.

In this study, we used a MRP-like approach to provide population-level estimates based on the 2017 NLA dissolved gas survey data. The remainder of this document provides the workflow, data, models, and code used to make these estimates. To begin, a number of Bayesian multilevel regression models were fit to the sample data. These models were fit and checked against both the data used to construct the models as well as hold-out data from future visits. After evaluating a number of different models, a final model was selected for use in inference. Because total N2O flux estimates needed to be estimated based on individual, waterbody-level measures (_e.g._, surface area), predictions from the model were made to each individual lake in the POI (n = 465,897). These predictions were assumed relevant to average conditions during the biological index period for each lake in 2017.

The objective of the modeling effort was to provide estimates for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of under-saturated water bodies (i.e., saturation ratio < 1) for the National-level POI and a number of sub-populations therein (_e.g._, ecoregion, state, size category). The gas estimates were also used to estimate the total flux of N2O gas attributable to the target populations. Saturation ratio was calculated as the ratio of dissolved to equilibrium N2O. Because dissolved and equilibrium N2O were observed on the same sample units (sites in lakes), the regression models were fit to their joint distribution and the ratio estimates were assembled as a derived quantity calculated from the joint posterior predictive distribution of those variables. The response variable was, therefore, modeled as multivariate in order to account for any potential statistical dependencies between dissolved and equilibrium N2O due to, for example, common dependencies on geography. Although predictions of the mean marginal probabilities from separate models may have provided comparable estimates, a joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. All of the models fit were constructed using the `brms` package [@Burkner_2017] in `R` [@R_Core_Team_2021] as an interface to Stan, a software package for fitting fully Bayesian models via Hamiltonian Monte Carlo [HMC, @Stan_Development_Team_2018_a; @Stan_Development_Team_2018_b; @Stan_Development_Team_2018_c]. More specific details on the model structures is provided throughout the "Model fitting" section below.

# Data
As explained in a previous data munging document document
(https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dgIndicatorAnalysis.html), duplicate dissolved gas samples were collected at a depth of ~0.1m at designated index sites distributed across 1091 lakes nationwide, of which 95 were sampled twice as repeat visits. This subset of revisit sites was used as a test set for assessing model fit and out-of-sample performance.

Water samples were analyzed via gas chromotography and concentrations were recorded to the nearest 0.001 nmol/L. As part of the survey design, each gas observation was indexed to an individual lake selected with unequal probability from 5 different lake size categories, $j \in j=1,...,J = 5$, according to surface area (ha), and from within a state, $k \in k=1,...,K = 48$, situated within an aggregated, WSA9 or Omernik ecoregion, $l \in l=1,...,L = 9$. All WSA9 ecoregions were represented, including the Xeric (XER), Western Mountain (WMT),  Northern Plains (NPL), Southern Plains (SPL), Temperate Plains (TPL), Coastal Plains (CPL), Upper Midwest (UMW), Northern Appalachian (NAP), and Southern Appalachian (SAP) regions. Gas data from the first and second (follow-up) visits were separated, with $n=984$ _vs._ $n=95$ observations each. In addition to the design variables, a number of potentially useful covariates were also indexed to the gas observations, including measures of NO3, surface water temperature, elevation, chlorophyll a content, dissolved oxygen content, and waterbody size (hectares). These gas data and covariates were previously described and munged at: https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dataMunge.html.

## Import
That originally munged gas dataset was imported below.

>>>>

---
title: "Modeling workflow: NLA N2O survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
    #  url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

# Load model and data objects from drive for knitting (if they already exist)
## vector of object names
objs <- c("dg",
          "df_model",
          "df_test",
          "sframe",
          "pframe",
          "samp_props",
          "n2o_mod1",
          "n2o_mod2",
          "n2o_mod3",
          "n2o_mod4",
          "n2o_mod5",
          "n2o_mod6",
          #"all_predictions_b",
          "n2o_survey_ests",
          "n2o_survey_ests_size",
          "sat_survey_ests",
          "sat_survey_ests_size"
          )

N <- length(objs)

# list of file paths for objects saved
f <- list()
for(i in 1:N){
    f[i] <- paste0(Sys.getenv("RENV_PROJECT"), "/modelFiles/",
    objs[i], ".rda")
    }

# load files if they exist
for(i in 1:N){
  if(file.exists(f[[i]]))
    load(f[[i]])
}

# clean up workspace
rm(f, i, objs, N)
```

# Introduction

This supplement documents the modeling workflow used for estimating
dissolved and equilibrium N2O gas concentrations and saturation ratios
for freshwater lakes (natural and man-made) across the continental
United States (CONUS). Data for the estimates came from the 2017 Nation
Lakes Assessment (NLA) survey, which aimed to provide representative
samples for describing environmental conditions across CONUS lakes
during the summer index period of 2017 [@USEPA_NLA_DESIGN_2017]. The
population of interest (POI) for this study was defined as all lakes and
reservoirs in the CONUS that were (1) larger than 1 hectare, (2) with at
least 0.1 hectares of open water, (3) at least 1 meter deep, and (4)
with a residence time of less than 1 week. The 2017 NLA sampling frame
consisted of 465,897 lakes delineated from georeferenced data documented
in the National Hydrography Dataset (NHD). The samples were drawn
according to a spatially balanced, stratified, and unequal probability
design. Sampling was stratified among categories of lake size (surface
area in hectares), WSA9 ecoregion, and US state (excluding AK and HI).
The probability of selection was greater for larger lakes relative to
smaller ones in order to sufficiently represent larger lakes in a sample
of practical size. Small lakes vastly outnumber large ones in the POI.

The POI was defined in this study explicitly by the NLA 2017 sampling
frame ([@USEPA_NLA_DESIGN_2017]). Therefore, the effective and stated
POI may depart in important ways that would be difficult to confirm in
practice. For example, it would be impractical to visit all lakes in the
sampling frame to determine which contained less than 0.1 ha of open
water or were less than 1 meter deep. These conditions were evaluated in
the field for sites that were sampled. If a field crew arrived at a site
that did not meet the conditions for inclusion, another site could be
chosen from an oversample frame. In effect, the size of the stated POI
is likely smaller than the 465,897 lakes described in the sampling frame
and, therefore, the possibility of some remaining biases in these
estimates should be considered, if only to emphasize that there remains
uncertainty in the population-level estimates beyond what was estimated
from modeling. Though it may be practically impossible to determine the
precise magnitude or direction in which any particular estimate could be
biased, it is clear that estimates could vary systematically given any
substantial mismatches between the POI and the sampling frame.

When making inferences to a POI from a sample, estimates should be
adjusted to address potential biases resulting from known differences
between the POI and sample. For this study, lakes in the POI were
missing from the sample not completely at random, but systematically and
according to the probability assigned by the survey design. This type of
"missingness" is not ignorable for making inferences from a sample, but
may be considered ignorable if the inferences are adjusted such that
they are conditional on the survey design [@Gelman_etal_2014, Ch. 8;
@Gelman_etal_2020 Ch. 17]. A straightforward approach to adjustment is
to simply include the design variables as predictors in a regression
model for the surveyed responses of interest [@Gelman_etal_2020, Ch.
17]. Then, in a second step referred to as "poststratification"
[@Gelman_etal_2020, Ch. 17], the model parameter estimates can be used
to make population-level estimates by summarizing over the known
distribution of the design variables in the POI or sub-populations
therein. Multilevel regression models are often preferred in this
approach, which is why it is often referred to as multilevel regression
with poststratification, or MRP [@Park_etal_2004; @Gelman_Little_1997;
@Gelman_etal_2020, Ch. 17; @Kennedy_Gelman_2021]. Multilevel regression
models are recommended because they naturally provide regularized
estimates along the design groupings, which can improve out-of-sample
inferences [@Gelman_Little_1997]. Likewise, estimates for group levels
that may be missing from the sample (e.g., an unsampled U.S. state), but
are part of the POI, are straightforward using the multilevel approach
[@Gelman_etal_2020 Ch. 17; @McElreath_2020]. For a worked and narrated
example using this approach, see a recent paper by Kennedy and Gelman
[@Kennedy_Gelman_2021]. For an example in the context of national
surveys of environmental resources, see a recent paper from Zachmann and
others [@Zachmann_etal_2022].

In this study, we used the 2017 NLA survey data and an adjustment
approach similar to MRP to provide population-level estimates for (1)
dissolved and equilibrium N2O concentrations; (2) the N2O saturation
ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of
under-saturated water bodies (i.e., saturation ratio \< 1). The
resulting estimates were later combined with lake-level attributes
(e.g., surface area) to provide estimates of N2O emission rates and
flux. A number of Bayesian multilevel regression models were fit to the
survey data and predictions from those models were checked against both
the data used in their fitting as well as a subset of held-out data from
re-visited sites. Because dissolved and equilibrium N2O were each
observed on the same sample units (a single site within each waterbody),
our regression models were constructed to estimate their joint
probability distribution. The multivariate response allowed for
conditional dependencies between dissolved and equilibrium N2O due to,
for example, physiography. Although predictions from separate models for
each N2O response may have provided comparable estimates, a joint model
was expected to better capture uncertainty and potentially improve
out-of-sample predictions, should the data be conditionally correlated
[@Warton_etal_2015; @Poggiato_etal_2021]. The saturation ratio
(saturation = dissolved N2O/equilibrium N2O) was estimated as a derived
quantity calculated from the joint posterior predictive distribution of
the two N2O variables.

Our initial regression models were focused on conditioning the N2O data
on the survey design factors only, which included multilevel groupings
of the sampled lakes according to WSA9 ecoregion, states within
ecoregions, and size classes within states. Although these initial
models performed well at replicating the observed group-level means,
further model checking indicated that they did a relatively poor job of
replicating some tail properties, such as the long right tail of the
observed dissolved N2O data. As a result of this misfit, the predictive
distribution for the saturation ratio was also consistently biased low
relative to the observed data. In subsequent models, we found that
including an NO3 covariate was particularly helpful for better
replicating these extreme values. The addition of predictors for surface
temperature and lake surface area, along with interaction terms for
these predictors with NO3, also lead to apparent improvements. Likewise,
the inclusion of surface temperature and elevation covariates in the
equilibrium N2O response resulted in improvements over the initial
models, as did added flexibility in the distributional terms
(heterogeneous errors) for both dissolved and equilibrium N2O.

A number of additional, potentially relevant covariates were evaluated
in exploratory analyses and models, but are not covered in this
document. These variables are, however, available to the public should
the reader wish to further explore their utility. These measures, which
were taken at the time of N2O sampling or constructed from those
measures, included chlorophyll a concentration, dissolved oxygen (DO)
content at the surface, and a buoyancy metric. Some of these variables
are also covered in the data munging document at:
<https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dgIndicatorAnalysis.html>.
An exploratory analysis using classification trees in that document
suggested that NO3 as well as both mean depth and chlorophyll a
concentration were associated with dissolved N2O concentration. The
decision to not include these variables in the final model was based on
several issues. First, our primary goal was to make population-level
estimates for N2O while considering the survey design structure. As a
result, we were less concerned with exploring the relative contributions
of environmental correlates to N2O variation. However, our initial
models excluding all covariates proved to result in relatively poor fits
to important aspects of the observed data. In an attempt to improving
the fit, the NO3 covariate was then included due to the relatively large
effects implied in exploratory analyses (see "Sample Data Exploration"
to follow). Conversely, our evaluations of exploratory plots and models
with chlorophyll a, surface DO, and the BF covariates did not result in
similarly substantial improvements. Second, these other covariates were
only available in the sample and, therefore, were not fully observed in
the POI. As such, using them to make population-level predictions would
have come with additional complexity (discussed below) that didn't
appear to be reciprocated by critical improvements in model fit.
Finally, several observations were missing from the sample for these
covariates. To include them in our model would have required sub-models
for the missing data, resulting in added complexity and computational
burden that, again, did not appear to be worth the added complexity and
computational burden. Hence, the progression of models presented in this
supplement was intended to efficiently convey the improvements in fit
following the addition of key covariates and parameterizations that
ultimately comprised the final model.

The first regression model we fit was constructed to estimate the joint
distribution of log-transformed dissolved and equilibrium N2O
conditional only on the the survey design factors, such that:

$$Y \sim MVN(\nu, \Sigma)$$

Indexing to each response, $p$:

$$Y = [y_{p=1}, y_{p=2}] \sim MVN(\nu = [\mu_{p=1}, \mu_{p=2}], \Sigma)$$

Indexing to the observation level, $i$, within each response:

$$\mu_{pijkl} = \alpha_{0(pi)} + \alpha_{1(pij)} + \alpha_{2(pijk)} + \alpha_{3(pijkl)}$$

$$\alpha_1 \sim MVN(0, \Lambda_1)$$ 
$$\alpha_2 \sim MVN(0, \Lambda_2)$$
$$\alpha_3 \sim MVN(0, \Lambda_3)$$

$$\Lambda_{\alpha = 1,..,3} = \begin{pmatrix} 1 & \tau ^2_{p=1} \\ \tau^2_{p=2} & 1 \end{pmatrix} \chi_{\alpha = 1,..,3} \begin{pmatrix} 1 & \tau^2_{p=1} \\ \tau^2_{p=2} & 1 \end{pmatrix}$$

$$\chi_{\alpha = 1,..,3} = \begin{pmatrix} 1 & \varrho_{p1,p2} \\ \varrho_{p1,p2} & 1 \end{pmatrix}$$

$$\Sigma = \begin{pmatrix} 1 & \sigma^2_{p=1} \\ \sigma^2_{p=2} & 1 \end{pmatrix} \Omega \begin{pmatrix} 1 & \sigma^2_{p=1} \\ \sigma^2_{p=2} & 1 \end{pmatrix}$$
$$\Omega = \begin{pmatrix} 1 & \rho_{p1,p2} \\ \rho_{p1,p2} & 1 \end{pmatrix}$$

Each log-transformed observation, $i \in 1,..,N=984$, for each N2O
response variable, $p \in 1:P=2$, was assumed to be drawn from a
multivariate normal (MVN) distribution with the parameters $\nu$ and
$\Sigma$, where $\nu$ is the multivariate mean and $\Sigma$ is a
covariance matrix containing the observation-level variances and
(residual) correlations.

The multivariate mean was defined by a vector of location parameters,
$\nu = [\mu_{p=1}, \mu_{p=2}]$, for each response, $p$. As noted above,
each location parameter was also defined according to a linear
combination of parameters indexed to each response $p$ and observation
$i$. This linear combination included a fixed global intercept,
$\alpha_0$, conditioned on the data, and three latent group-level
effects matrices, $\alpha_1, \alpha_2, \alpha_3$. The effects in each of
these matrices were also drawn from MVN probability distributions
centered on zero in $p$-dimensional space. The spread of the effects
around zero were determined by a corresponding variance-covariance
matrix, $\Lambda_1, \Lambda_2, \text{or } \Lambda_3$. The standard
deviation parameters, $\tau$ , of those matrices were conditioned on the
data and captured group-level standard deviations that controlled the
spread of the group-level effects for each response. The $\chi$
parameter contained the group-level residual correlation matrix, wherein
$\varrho$ denotes the group-level residual correlations (i.e., between
responses).

The explicit indexing in the notation above conveys the relationship
between the parameters and each observation, $i$, and emphasizes the
nested structure of the observations within the group effects.
Specifically, each observation, $i$, was nested in a lake size category,
$l$, which was nested in a state, $k$, and ecoregion, $j$. The
group-level effects in $\alpha_1$, therefore, accounted for
ecoregion-level deviations from the global mean; $\alpha_2$ accounted
for state-level group effects nested in ecoregions; and $\alpha_3$
accounted for lake size category effects within states and ecoregions.

Finally, the observation-level covariance term, $\Sigma$, comprised of
the observation-level standard deviations, $\sigma$, for each response,
and $\Omega$ or the observation-level residual correlation matrix,
wherein $\rho$ captured residual correlation between responses.

For model fitting, priors were needed for all parameters conditioned
directly on the data, which included the global intercepts, $\alpha_0$;
the scale parameters of the latent group effects, $\tau$ and $\sigma$;
and the correlation parameters, $\varrho$ and $\rho$. A normal or
Gaussian prior, $N(\mu = 2, \sigma = 1)$ centered near the (log-scale)
data means, was used as the prior for the global mean parameter for each
response. This prior was considered minimally informative as it placed
most (\~80%) of the prior mass over values between about 2 and 27 ng/L
for median N2O and median N2O equilibrium concentration; and included
support in the tails for values approaching 0 ng/L on the lower end and
80 ng/L on the high end. We placed $Exp(2)$ priors over all scale
parameters, which placed most of the support between values very close
to 0 and values near 1 (central 80% density interval from approximately
0.005 to 1.15). Finally, for the correlation matrices, an $LKJ(\eta =2)$
prior was used, which, for a 2-dimensional response, placed most support
for correlations between approximately -0.9 and 0.9. This prior seemed
reasonable as there is no direct causal link between these responses
that would warrant a very strong correlation . Any residual dependence
was expected to be conditional or indirect due to, for example, a common
correlate (e.g., elevation, temperature). For more information on prior
choice recommendations in Stan, see:
<https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations>

The $\textbf{brms}$ package [@Burkner_2017] for $\textbf{R}$
[@R_Core_Team_2021] was used to fit all of the models in a fully
Bayesian setting. The formula syntax of the $\textbf{brms}$ package is
similar to the syntax used in the $\textbf{lme4}$ package that is widely
used to fit mixed effects models in frequentist settings. In either
package, the linear predictor for $\mu$ described above, but ignoring
multivariate dependencies, would be expressed as:

In the $\textbf{brms}$ package, there is additional functionality and
formula syntax for multivariate responses, which allows for the varying
intercepts of the separate linear predictors in a multivariate model to
be correlated across responses, e.g.,:

The above formula syntax indicates that the linear predictor for both
N2O responses have the same group-level varying effects terms, and that
the effects associated with each of those terms can be correlated across
responses. The relevant sections of $\textbf{R}$ code that are helpful
for describing the first model above using $\textbf{brms}$ syntax are
below:

```{r model_1_text, eval=FALSE, include=TRUE}
# Assign separate linear predictors for the dissolved and equilibrium N2O responses
bf_n2o <- bf(log(n2o) ~ 1 + 
               (1 | a | WSA9) + 
               (1 | b | WSA9:state) + 
               (1 | c | WSA9:state:size_cat),
             family = gaussian())

# Set priors for parameters
priors <- c(
  prior(normal(2, 1), class = "Intercept", resp = "logn2o"), # alpha_0
  prior(exponential(2), class = "sd", resp = "logn2o"), # sd alpha_1..3
  prior(exponential(2), class = "sigma", resp = "logn2o"), # observation level sd
  prior(normal(2, 1), class = "Intercept", resp = "logn2oeq"), 
  prior(exponential(2), class = "sd", resp = "logn2oeq"),
  prior(exponential(2), class = "sigma", resp = "logn2oeq"),
  prior(lkj(2), class = "rescor"), # observation level residual correlation
  prior(lkj(2), class = "cor") # correlation for group effects
  )

# Fit the multivariate model
n2o_mod1 <- brm(bf_n2o + bf_n2oeq + set_rescor(rescor = TRUE), ...)
```

The first two objects assigned in the above code example are the linear
predictors assigned to the dissolved and equilibrium N2O responses. The
second object is a list of priors as defined above. The third object
results from the `brm()` call to fit the model. In this syntax, the two
linear predictor objects are "added" along with the
`set_rescor(rescor = TRUE)` option, which defines whether the residual
correlation is to be modeled (TRUE) or not (FALSE). Note that, with
`set_rescor(rescor = FALSE)` and without the correlated group effects
(i.e., 'a', 'b', 'c'), $\textbf{brms}$ would effectively treat the two
responses as entirely independent in fitting.

For the remainder of this document, only this simplified syntax is
presented to describe the model structure. For more information on
$\textbf{brms}$ functionality and syntax with multivariate response
models, see the package vignette:
<https://cran.r-project.org/web/packages/brms/vignettes/brms_multivariate.html>.

The relevant code for the final model used for inference and described
in the main paper is below:

```{r model_final_text, eval=FALSE, include=TRUE}
# Assign separate linear predictors for each response: N2O, N2o-eq, Surface temp, and NO3 level.

# N2O
bf_n2o <- bf(n2o ~ mo(no3_cat) +
               log_area +
               surftemp + 
               mo(no3_cat):log_area +
               mo(no3_cat):surftemp +
               (mo(no3_cat) | a | WSA9) + 
               (mo(no3_cat) | b | WSA9:state) + 
               (1 | c | WSA9:state:size_cat),
             shape ~ log_area +
               mo(no3_cat) +
               (1 | WSA9) + 
               (1 | WSA9:state) + 
               (1 | WSA9:state:size_cat),
             family = Gamma(link = "log"))

# Equilibrium N2O
bf_n2oeq <- bf(n2o_eq ~ surftemp +
                 log_elev +
                 surftemp:log_elev +
                 (1 | a | WSA9) + 
                 (1 | b | WSA9:state) +
                 (1 | c | WSA9:state:size_cat),
             shape ~ surftemp +
               log_elev +
               (1 | WSA9) + 
               (1 | WSA9:state) + 
               (1 | WSA9:state:size_cat),
             family = Gamma(link = "log"))

# Surface temperature
bf_temp <- bf(surftemp ~ lat +
                s(log_elev) +
                s(jdate) +
                (1 | a | WSA9) + 
                (1 | b | WSA9:state) +
                (1 | c | WSA9:state:size_cat),
              shape ~ lat,
              family = Gamma(link = "log"))

# NO3 level
bf_no3 <- bf(no3_cat ~ surftemp +
               log_area +
               (1 | a | WSA9) +
               (1 | b | WSA9:state) +
               (1 | c | WSA9:state:size_cat),
             family = cumulative(link = "logit", threshold="flexible"))

# Assign priors 
priors <- c( # N2O
  prior(normal(2, 1), class = "Intercept", resp = "n2o"),
  prior(normal(0, 1), class = "b", resp = "n2o"),
  prior(exponential(2), class = "sd", resp = "n2o"),
  prior(normal(5, 4), class = "Intercept", dpar = "shape", resp = "n2o"),
  prior(normal(0, 1), class = "b", dpar = "shape", resp = "n2o"),
  prior(exponential(2), class = "sd", dpar = "shape", resp = "n2o"),

  # Equil N2O
  prior(normal(2, 1), class = "Intercept", resp = "n2oeq"), 
  prior(normal(0, 1), class = "b", resp = "n2oeq"),  
  prior(exponential(2), class = "sd", resp = "n2oeq"),
  prior(normal(5, 4), class = "Intercept", dpar = "shape", resp = "n2oeq"),
  prior(normal(0, 1), class = "b", dpar = "shape", resp = "n2oeq"),
  prior(exponential(2), class = "sd", dpar = "shape", resp = "n2oeq"),

  # Surface temp
  prior(normal(3, 1), class = "Intercept", resp = "surftemp"), 
  prior(normal(0, 1), class = "b", resp = "surftemp"), 
  prior(exponential(0.5), class = "sds", resp = "surftemp"),
  prior(exponential(2), class = "sd", resp = "surftemp"),
  prior(normal(5, 4), class = "Intercept", dpar = "shape", resp = "surftemp"),
  prior(normal(0, 1), class = "b", dpar = "shape", resp = "surftemp"),

  # NO3
  prior(normal(0, 3), class = "Intercept", resp = "no3cat"),
  prior(normal(0, 1), class = "b", resp = "no3cat"),
  prior(exponential(1), class = "sd", resp = "no3cat"),

# Fit the multivariate model
n2o_mod6 <- brm(bf_n2o + bf_n2oeq + bf_temp + bf_no3 + set_rescor(rescor = FALSE), ...)
```

In this final model, the dissolved and equilibrium N2O responses were
each modeled with Gamma distributed errors. The Gamma error structure
appeared to result in slightly better performance in the predictive
checks compared to the Gaussian errors. This was primarily apparent in
the saturation ratio checks, which may have been more sensitive to model
performance in the tails of the N2O responses. Others have indicated
that the Gamma error distribution can work well for modeling dissolved
N2O data [@Webb_etal_2019]. The linear predictor for dissolved N2O
included the survey design effects along with covariate terms for lake
area (ha), NO3 level, and surface water temperature. To accomodate
heterogeneous errors, the shape parameter of the Gamma distribution was
modeled as a function of NO3 level, lake area, and the survey design
factors. The equilibrium N2O response was modeled as a function of the
survey design factors, lake elevation, and lake surface temperature. The
shape parameter was, again, allowed to vary conditional on the survey
factors, surface temperature, and elevation.

The sub-model for the surface temperature response also assumed a Gamma
distributed error distribution and covariates included the survey design
variables, latitude, elevation, and Julian date. The shape parameter was
also modeled as a function of latitude to address increasing temperature
variance along the latitudinal gradient.

The NO3 sub-model was a cumulative logit model and the linear predictor
included all of the survey factors as well as surface temperature and
lake area.

Note that there was no observation-level residual correlation term for
this final model, since the residuals are undefined for the Gamma and
cumulative logit models. Dropping the observation-level residual
correlation term was deemed a reasonable compromise that enabled the
inclusion of NO3 as a covariate on dissolved N2O. The random intercepts,
however, still allowed for potential correlations between the four
responses at the group levels. As mentioned previously, including
covariates that were not available in the full POI necessitated some
additional steps in the poststratification process. To make
population-level inferences from a model with covariates, those
covariates need to be (1) fully observed across the POI or (2) their
"missingness" needs to be addressed with modeling. For the lake area and
elevation covariates, for example, data was available for the entire POI
from previously compiled geospatial databases. However, neither surface
temperature or NO3 were observed for lakes outside of the sample. That
is, they were partially observed with respect to the POI. Therefore, the
final multivariate model, coded above included these as additional
responses, with each also being conditioned on the survey design
structure and the fully observed covariates. The dependence structure
among the fully and partially observed variables in this model could be
described such that:

Variables in bold text above were partially observed with respect to the
POI (i.e., observed only in the sample), whereas variables not in bold
were considered fully observed. The partially observed variables, being
dissolved and equilibrium N2O, NO3, and surface temperature, were each
modeled conditional on the survey design variables and other partially
and/or fully observed covariates. This piece-wise approach then required
a more complex set of post-processing steps compared to a typical MRP
analysis. In order to propagate uncertainty through to the
population-level inferences, the fitted model was first used to predict
surface temperature in the target population, since it depended only on
the fully observed covariates. That predictive distribution was then
used alongside the relevant fully observed covariates to predict NO3 in
the target population. Finally, these predictive distributions for
temperature and NO3 were used to predict the N2O responses. These steps
were carried out in the "Predict to population" section below. This
approach is similar to MRP approaches with "non-census" variables
[@Kastellec_etal_2015]. It could also be described as a form of Bayesian
structural equation model [@Merkle_etal_2021; @Merkle_Rosseel_2018].

Finally, because N2O flux was to be later estimated as a function of
lake-level measures (*e.g.*, surface area), predictions for N2O were
made to each of the 465,897 individual lakes in the POI. For population
level estimates of dissolved N2O, equilibrium N2O, and N2O saturation,
the lake-level predictions were summarized across the relevant units
(*e.g.*, CONUS, ecoregion). Predictions were assumed relevant for
average conditions during the biological index period of 2017.
